{Simulating the Brownian Dynamics of a Two-Dimensional Model for the Dynein Motor Protein}\vspace{1.5em}\\ \includegraphics[width=0.5\columnwidth]{OSU_logo.png}\vspace{0.5em}\\ \large{Oregon State University}

### 2019-02-22 papers/waczak-thesis/thesis/chapters/introduction.tex
thesis updates
thesis update

\section{Cellular Processes}
	\subsection{The Cytoskeleton}
	\subsection{Mitosis} 
\section{Motor Proteins}

>>>>

Cells are the fundamental unit for all life on Earth. In order
for them to perform their biological tasks, the complex structures they contain
must be able to replicate and move within their aqueous environment. This
transport problem  poses a unique challenge as diffusion alone is far to slow,
and random, a process to be able to produce movement on cellular timescales[].
Nature's solution to this problem is twofold. Within each cell there is a vast network of interweaving filaments called the \textit{cytoskeleton} that spans the gaps between the substructures of the cell. This cytoskeleton serves as a highway to a class of proteins called \textit{molecular motors} which can be distinguished into three types: kinesin, myosin, and dynein. Molecular motors attach to cellular cargo and drag it along the cytoskeleton by hydrolyzing Adenosine triphosphate (ATP) in order to convert stored chemical energy into mechanical work\cite{alberts_molecular_2002}. Correct operation of these motors results in processes such as cell division, muscle contraction, and axonal transport; failure can lead to debilitating diseases such as ALS \cite{boillee_als:_2006}.

\begin{figure}[!hbt]
	\centering 
	\includegraphics[width=0.5\columnwidth]{motor_comparison}
	\caption{Comparison of the three primary motor proteins (from left to right): myosin, kinesin, dynein. Image taken from \cite{philips__nodate}.}
	\label{fig:motor proteins} 
\end{figure}

Astonishingly, each of these motors is composed of two legs that combine to form a bipedal structure (see figure \ref{fig:motor proteins}). that steps like a human as it moves along the cytoskeleton. Myosin and kinesin behave like a normal person; their steps alternate between opposing domains. This motion is called \textit{hand over hand} stepping \cite{noauthor_drunk_nodate}. Dynein, however, is unique in its motion as it takes hand-over-hand steps but also \textit{inchworms} meaning that it often steps twice with the same foot. Furthermore, dynein even takes frequent backwards steps as shown in the stepping histogram in figure \ref{fig:weihong histogram} \cite{qiu_dynein_2012}. Collectively this behavior has earned dynein the title of \textit{drunken walker} which compactly summarizes its peculiar tendencies \cite{noauthor_drunk_nodate}. \\

\begin{figure}[!hbt]
	\centering
	\includegraphics[width=0.5\columnwidth]{weihong_histogram}
	\caption{Stepping histogram showing the distribution of step sizes for \textit{in vitro} dynein. 1-D data corresponds to the projection of the step along the axis of the microtubule. 2-D data corresponds to the total step size over the two-dimensional structure. Figure taken form \cite{qiu_dynein_2012}.}  
	\label{fig:weihong histogram} 
\end{figure}

This unique stepping behavior makes dynein an interesting subject to study. There have been significant efforts aimed at characterizing the complex, multi-phase chemical cycle involved in the hydrolysis of ATP\cite{cianfrocco_mechanism_2015}. However, the dynamics of dynein's physical motion through space remains an open question. In this thesis, we examine a simplified physical model for the motion of dynein using Brownian dynamics. By simulating this model, we demonstrate it's ability to replicate of many observed properties of the protein. We will also use the model to comment a proposed mechanism for how dynein generates most of its movement, namely the powerstroke model.

### 2019-02-22 papers/waczak-thesis/thesis/chapters/theory.tex
thesis update

Now that I have established some facts about motor proteins as well as specifics about dynein's many odd behaviors, we may begin our discussion of how to construct and test a physical model for its motion. First, we note that dynein's large mass of roughly 2MDa \cite{johnson_structure_1983} as compared to similar motor proteins such as Kinesin (a few hundred kDa) \cite{liao_kinesin_1998} exaggerates the effect of interactions with the  aqueous environment of the cell. This results in a much more significant drag force as well as a random element from water molecule collisions that changes the overall dynamics of the system. The set of equations governing this motion is called Brownian dynamics and is what we use to simulate the trajectory of dynein along the microtubule. 
	\section{Brownian dynamics}
	To model dynein we treat its various domains (binding, motor, and tail) as spheres held together by massless rigid-rods. In general, the dynamics of a constant mass particle are governed by Newton's second law
	\begin{equation}
		\sum_i \mathbf{F}_i = \dot{\mathbf{p}} = m\ddot{\mathbf{x}}
	\end{equation}
	Thus, determining the dynamics of a system boils down to describing the various forces which act upon it and then solving the resulting differential equation for some set of initial conditions. For particles in a solvent, there are three important forces to consider. The first is linear drag, which we write as 
	\begin{equation}
		F_\text{drag} = -\gamma\dot{\mathbf{x}}
	\end{equation}
	so that the force is proportional to the velocity by a factor $\gamma$s. For a spherical particle of radius $a$ moving through a solvent with liquid viscosity $\eta$ this was derived by Stoke to be 
	\begin{equation}
		\gamma = 6\pi\eta a
	\end{equation}

@@

	If we assume further that $|\gamma\dot{\mathbf{x}}|\gg |m\ddot{\mathbf{x}}|$, or in other words, we take the limit of strong drag, then we arrive at the definition of Brownian Dynamics. This allows us to neglect the acceleration term resulting in the following differential equation:
	\begin{equation}
		\dot{\mathbf{x}} = \frac{1}{\gamma}\left( \mathbf{F}_\text{interact} + \mathbf{F}_\text{rand} \right)
	\end{equation}
	We must now solve the above for equations of motion.

	\section{A Simplified Model for Dynein}
	As mentioned, we model dynein as a system of spherical domains held together by massless rigid-rods with the whole system moving according to Brownian dynamics. The random forces on each domain are responsible for diffusing the system and creating the expected chaotic motion, but in order to have dynein prefer one side of the microtubule on average (as is the case for real dynein) we need to impose some \textit{extra} structure on the protein.

	\begin{figure}[hbt!]
		\centering
		\includegraphics[angle=-90, width=0.6\columnwidth]{torque_angles}
		\caption{Example of dynein in the onebound configuration with torque angles $\{\phi_i\}$ shown. \textbf{NOTE: double check w/ David on how we select the angles e.g. do we take min of the two possible angles? How do we define the equilibrium angles?}}
		\label{fig:torque angles}
	\end{figure}

	To do this we steer each domain towards an equilibrium configuration by a Hooke's law restorative torque with magnitude of the following form
	\begin{equation}
		\Gamma_i = -c_i\left(\varphi_i-\varphi_{i,\text{eq}}\right)
	\end{equation}
	Here the $c_i$ are spring constants (a free parameter of the model) and the set $\{\varphi_i\}$ are the angles measured between the $i-1$ and $i+1$ domains. Perhaps more useful is the harmonic energy associated with each torque given as 
	\begin{equation}
		U_i = \frac{1}{2}c_i(\varphi_i-\varphi_{i,\text{eq}})^2
	\end{equation} 
	Note that for binding domains this angle may be measured to the microtubule or not measured if the domain is unbound from the microtubule. The angles $\{\varphi_{i, \text{eq}}\}$ are the preferred angles of the system as illustrated in figure \ref{fig:torque angles}. Torques are related to forces by 
	\begin{equation}
		\mathbf{\Gamma} = \mathbf{r}\times\mathbf{F}
	\end{equation}
	so that knowing the distance between each domains means that we have specified each of the $\mathbf{\Gamma}_i$ and $\mathbf{r}_i$. We desire to determine $\mathbf{F}_i$ from this information however there is no unique way to invert a cross product. Therefore we must specify an algorithm for determining the forces that result from each torque.

	\section{Equations of Motion}
	To model how dynein steps we introduce two states hereafter referred to as the one bound state and the both bound state. In the one bound state, all but one of the domains are free to diffuse whilst the last is attached to the microtubule. In the both bound state, both binding domains are attached. The slight difference in geometry between the two scenarios leads to two sets of equations for motion that differ slightly. The following section details their derivation.

@@

		\section{The both bound state}
		\section{Calculating Forces and Torques}
			Here I will discuss in detail how we calculate the forces and show how this is done in code 
		\section{Determining constants of the model}
		The stalk and tail lengths $\ell_S, \ell_T$ as well as a set of equilibrium angles $\{\varphi_{i,\text{eq}}\}$ were determined by examining microscopy images of dynein in various configurations. 
		\begin{figure}[hbt!]
			\centering
			\includegraphics[width=0.3\columnwidth]{schematic-2-superimposed}
			\caption{Our model superimposed over microscopy image of Dynein \textbf{NOTE: add source} for determining lengths and equilibrium angles} 
			\label{fig:superimpmosed}
		\end{figure}
	\section{The Code}
		\subsection{The Random Force}

>>>>

Now that I have established some facts about motor proteins as well as specifics about dynein's many odd behaviors, we may begin our discussion of how to construct and test a physical model for its motion. First, we note that dynein's large mass of roughly 2MDa \cite{johnson_structure_1983} as compared to similar motor proteins such as Kinesin (a few hundred kDa) \cite{liao_kinesin_1998} exaggerates the effect of interactions with the  aqueous environment of the cell. This results in a much more significant drag force as well as a random element from water molecule collisions that changes the overall dynamics of the system. The set of equations governing this motion is called Brownian dynamics and is what we use to simulate the trajectory of dynein along the microtubule. 
	\section{Brownian dynamics}
	To model dynein we treat its various domains (binding, motor, and tail) as spheres held together by massless rigid-rods. In general, the dynamics of a constant mass particle are governed by Newton's second law
	\begin{equation}
		\sum_i \mathbf{F}_i = \dot{\mathbf{p}} = m\ddot{\mathbf{x}}
	\end{equation}
	Thus, determining the dynamics of a system boils down to describing the various forces which act upon it and then solving the resulting differential equation for some set of initial conditions. For small particles in a solvent, there are three important forces to consider. The first is linear drag, which we write as 
	\begin{equation}
		F_\text{drag} = -\gamma\dot{\mathbf{x}}
	\end{equation}
	so that the drag force is proportional to the velocity by a factor $\gamma$. For a spherical particle of radius $a$ moving through a solvent with liquid viscosity $\eta$ this was derived by Stoke to be 
	\begin{equation}
		\gamma = 6\pi\eta a
	\end{equation}

@@

	If we assume further that $|\gamma\dot{\mathbf{x}}|\gg |m\ddot{\mathbf{x}}|$, or in other words, we take the limit of strong drag, then we arrive at the definition of Brownian Dynamics. This allows us to neglect the acceleration term resulting in the following differential equation
	\begin{equation}
		\dot{\mathbf{x}} = \frac{1}{\gamma}\left( \mathbf{F}_\text{interact} + \mathbf{F}_\text{rand} \right)
	\end{equation}
	which we must now solve for the equations of motion of our model.

	\section{A Simplified Model for Dynein}
	As mentioned, we model dynein as a system of spherical domains held together by massless rigid-rods with the whole system moving according to Brownian dynamics. The random forces on each domain are responsible for diffusing the system and creating the expected chaotic motion, but in order to have dynein prefer walking towards one side of the microtubule, as is the case for real dynein, we need to impose some \textit{extra} structure on the protein.

	\begin{figure}[hbt!]
		\centering
		\includegraphics[angle=-90, width=0.6\columnwidth]{torque_angles}
		\caption{Example of dynein in the one bound configuration with torque angles $\{\varphi_i\}$ shown.}
		\label{fig:torque angles}
	\end{figure}

	To do this we steer each domain towards an equilibrium configuration by a Hooke's law restorative torque with magnitude of the following form
	\begin{equation}
		\Gamma_i = -c_i\left(\varphi_i-\varphi_{i,\text{eq}}\right)
	\end{equation}
	Here the $c_i$ are spring constants (a free parameter of the model) and the set $\{\varphi_i\}$ are the angles measured between the $i-1$ and $i+1$ domains. Equally useful is the harmonic energy associated with each torque. These are given by
	\begin{equation}
		U_i = \frac{1}{2}c_i(\varphi_i-\varphi_{i,\text{eq}})^2
	\end{equation}

	Real dynein is composed of a pair of identical heavy chains. Therefore, our angle convention is chosen starting at the $0^{th}$ domain in figure \ref{fig:torque angles}. The same convention is then repeated for the second leg with the following correspondence 
	\begin{align}
		\varphi_{0,eq} &\longleftrightarrow \varphi_{4,eq} \\ 
		\varphi_{1,eq} &\longleftrightarrow \varphi_{3,eq}
	\end{align}
	where the angles $\{\varphi_{i, \text{eq}}\}$ are the preferred angles of the system as illustrated in figure \ref{fig:torque angles}. The torques on each domain are related to forces by 
	\begin{equation}
		\mathbf{\Gamma} = \mathbf{r}\times\mathbf{F}
	\end{equation}
	so that knowing the distance between each domains means that we have specified each of the $\mathbf{\Gamma}_i$ and $\mathbf{r}_i$. In order to dynamically move each domain we must invert equation 4.10 to determine $\mathbf{F}_i$, however there is no unique way to invert such a cross product. Therefore, we must specify an algorithm for determining the forces that result from each torque.

	\section{Calculating Forces and Torques}
	Here I will discuss in detail how we calculate the forces and show how this is done in code

	\section{Equations of Motion}
	Now that we have specified the forces, the dynamics of each domain are determined by equation 4.5 so that in order to simulate the complicated structure we must solve a system of coupled first order differential equations. In doing so, it is convenient to introduce two states to distinguish between dynein's stepping and when both of its binding domains are attached to the microtubule. We shall refer to these configurations hereafter as the one bound state and the both bound state.

	In the one bound state, all but one of the domains are free to diffuse while the last is attached to the microtubule. In the both bound state, both binding domains are attached. This slight difference in geometry between the two scenarios leads to two sets of equations for motion that differ slightly. The following section details the derivation of these equations of motion.

@@

		\subsection{The both bound state}

		\section{Determining constants of the model}
		The stalk and tail lengths $\ell_S, \ell_T$ as well as a set of equilibrium angles $\{\varphi_{i,\text{eq}}\}$ were determined by examining microscopy images of dynein in various configurations. 
		\begin{figure}[hbt!]
			\centering
			\includegraphics[width=0.3\columnwidth]{schematic-2-superimposed}
			\caption{Our model superimposed over microscopy image of Dynein \textbf{NOTE: add source} for determining lengths and equilibrium angles} 
			\label{fig:superimpmosed}
		\end{figure}
	\section{The Code}
		\subsection{The Random Force}
